---
title: "Step 2 — Quality Control"
subtitle: "Jackson & Fischer et al. (2020) Breast Cancer | imcdatasets"
author: "mass-cytometry-explorations"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    fig-width: 9
    fig-height: 6
    embed-resources: true
conda:
  environment: mass-cytometry
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Overview

This notebook performs **Step 2** of the replication workflow: quality control (QC)
for images, cells, and markers before downstream preprocessing and clustering.

Inputs from Step 1:

- `results/sce_raw.rds`
- `results/acquisition_summary/cells_per_image.csv` (optional; recomputed here anyway)

Outputs from this notebook:

- `results/qc_summary/cells_per_image_qc.csv`
- `results/qc_summary/cell_qc_metrics.csv`
- `results/qc_summary/marker_qc_metrics.csv`
- `results/sce_qc_filtered.rds`

---

## Setup

```{r setup}
# Load required packages.
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(ggplot2)
  library(dplyr)
  library(patchwork)
})
# Desired working directory.
target_dir <- "/home/lizalfos/mass-cytometry-explorations/notebooks"

# Check current working directory and switch if needed.
if (!identical(normalizePath(getwd()), normalizePath(target_dir))) {
    setwd(target_dir)
}

# Confirm current working directory.
cat("Working directory:", getwd(), "\n")
```

```{r paths}
# Create correct file paths for first run.
dir.create(file.path("..", "results", "qc_summary"),
           recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("..", "figures", "02_quality_control"),
           recursive = TRUE, showWarnings = FALSE)
```

---

## Load data

```{r load-sce}
# Validating presence of data for quality control.
sce_path <- file.path("..", "results", "sce_raw.rds")
if (!file.exists(sce_path)) {
  stop("Missing results/sce_raw.rds. Please render notebooks/01_data_acquisition.qmd first.")
}

sce <- readRDS(sce_path)
sce
```

---

## QC thresholds

These thresholds are conservative defaults and can be tuned later if needed.

```{r thresholds}
# QC thresholds.
min_cells_per_image <- 100L
low_mad_multiplier_total <- 3
high_mad_multiplier_total <- 5
low_mad_multiplier_detected <- 3
low_prevalence_pct <- 1
```

---

## Image-level QC

```{r image-qc-metrics}
cells_per_image <- as.data.frame(colData(sce)) |>
  dplyr::count(image_name, name = "n_cells") |>
  dplyr::mutate(
    fail_image_qc = n_cells < min_cells_per_image
  )

keep_images <- cells_per_image |>
  dplyr::filter(!fail_image_qc) |>
  dplyr::pull(image_name)

cells_per_image |>
  dplyr::summarise(
    images_total = dplyr::n(),
    images_fail = sum(fail_image_qc),
    images_keep = sum(!fail_image_qc)
  )
```

```{r fig-image-qc}
#| fig-cap: "Cell counts per image. Red bars are image-level QC failures."
ggplot(cells_per_image,
       aes(x = reorder(image_name, n_cells), y = n_cells, fill = fail_image_qc)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = min_cells_per_image, linetype = "dashed", colour = "red") +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick")) +
  labs(
    title = "Image-level QC",
    subtitle = sprintf("Threshold: at least %d cells per image", min_cells_per_image),
    x = NULL, y = "Cell count", fill = "Fail QC"
  ) +
  theme_bw(base_size = 9) +
  theme(axis.text.y = element_text(size = 6))
```

---

## Cell-level QC

```{r cell-qc-metrics}
counts_mat <- assay(sce, "counts")

cell_total_counts <- colSums(counts_mat)
cell_detected_markers <- colSums(counts_mat > 0)

low_total_threshold <- median(cell_total_counts) -
  low_mad_multiplier_total * mad(cell_total_counts)
high_total_threshold <- median(cell_total_counts) +
  high_mad_multiplier_total * mad(cell_total_counts)
low_detected_threshold <- median(cell_detected_markers) -
  low_mad_multiplier_detected * mad(cell_detected_markers)

low_total_threshold <- max(low_total_threshold, 1)
low_detected_threshold <- max(low_detected_threshold, 1)

cell_qc <- data.frame(
  cell_id = colnames(sce),
  image_name = sce$image_name,
  total_counts = cell_total_counts,
  detected_markers = cell_detected_markers,
  fail_low_total = cell_total_counts < low_total_threshold,
  fail_high_total = cell_total_counts > high_total_threshold,
  fail_low_detected = cell_detected_markers < low_detected_threshold
) |>
  dplyr::mutate(
    fail_cell_qc = fail_low_total | fail_high_total | fail_low_detected
  )

cell_qc |>
  dplyr::summarise(
    cells_total = dplyr::n(),
    cells_fail = sum(fail_cell_qc),
    cells_keep = sum(!fail_cell_qc)
  )
```

```{r fig-cell-qc-1}
#| fig-cap: "Distribution of total ion counts per cell with low/high QC thresholds."
ggplot(cell_qc, aes(x = total_counts)) +
  geom_histogram(bins = 80, fill = "grey60", colour = "white") +
  geom_vline(xintercept = low_total_threshold, colour = "firebrick", linetype = "dashed") +
  geom_vline(xintercept = high_total_threshold, colour = "firebrick", linetype = "dashed") +
  labs(
    title = "Cell-level QC: total counts",
    subtitle = "Dashed lines = fail thresholds",
    x = "Total counts per cell", y = "Number of cells"
  ) +
  theme_bw()
```

```{r fig-cell-qc-2}
#| fig-cap: "Relationship between detected markers and total counts. Red points fail cell-level QC."
ggplot(cell_qc, aes(x = detected_markers, y = total_counts, colour = fail_cell_qc)) +
  geom_point(alpha = 0.35, size = 0.6) +
  scale_colour_manual(values = c("FALSE" = "grey40", "TRUE" = "firebrick")) +
  labs(
    title = "Cell-level QC flags",
    x = "Detected markers per cell", y = "Total counts per cell", colour = "Fail QC"
  ) +
  theme_bw()
```

---

## Marker-level QC

```{r marker-qc}
exprs_assay <- if ("exprs" %in% assayNames(sce)) assay(sce, "exprs") else counts_mat

marker_qc <- data.frame(
  marker_name = rownames(sce),
  mean_expr = rowMeans(exprs_assay),
  median_expr = apply(exprs_assay, 1, median),
  pct_cells_nonzero = rowMeans(counts_mat > 0) * 100
) |>
  dplyr::mutate(
    fail_low_prevalence = pct_cells_nonzero < low_prevalence_pct
  ) |>
  dplyr::arrange(pct_cells_nonzero)

marker_qc |>
  dplyr::summarise(
    markers_total = dplyr::n(),
    markers_low_prevalence = sum(fail_low_prevalence)
  )
```

```{r fig-marker-qc}
#| fig-cap: "Percent of cells with non-zero signal for each marker."
ggplot(marker_qc,
       aes(x = reorder(marker_name, pct_cells_nonzero),
           y = pct_cells_nonzero,
           fill = fail_low_prevalence)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = low_prevalence_pct, colour = "firebrick", linetype = "dashed") +
  coord_flip() +
  scale_fill_manual(values = c("FALSE" = "steelblue", "TRUE" = "firebrick")) +
  labs(
    title = "Marker-level QC",
    subtitle = sprintf("Low-prevalence threshold: %.1f%% of cells", low_prevalence_pct),
    x = NULL, y = "% cells with non-zero signal", fill = "Fail QC"
  ) +
  theme_bw(base_size = 9)
```

---

## Apply QC filters

```{r apply-qc}
keep_cells <- (!cell_qc$fail_cell_qc) & (cell_qc$image_name %in% keep_images)
sce_qc <- sce[, keep_cells]

cat(sprintf("Cells before QC: %d\n", ncol(sce)))
cat(sprintf("Cells after QC:  %d\n", ncol(sce_qc)))
cat(sprintf("Markers retained: %d\n", nrow(sce_qc)))
```

---

## Save outputs

```{r save-outputs}
write.csv(cells_per_image,
          file = file.path("..", "results", "qc_summary", "cells_per_image_qc.csv"),
          row.names = FALSE)

write.csv(cell_qc,
          file = file.path("..", "results", "qc_summary", "cell_qc_metrics.csv"),
          row.names = FALSE)

write.csv(marker_qc,
          file = file.path("..", "results", "qc_summary", "marker_qc_metrics.csv"),
          row.names = FALSE)

saveRDS(sce_qc, file = file.path("..", "results", "sce_qc_filtered.rds"))

cat("Saved:\n")
cat("  results/qc_summary/cells_per_image_qc.csv\n")
cat("  results/qc_summary/cell_qc_metrics.csv\n")
cat("  results/qc_summary/marker_qc_metrics.csv\n")
cat("  results/sce_qc_filtered.rds\n")
```

---

## Session information

```{r session-info}
sessionInfo()
```

---

::: {.callout-tip}
## Next step

Proceed to **[Step 3 — Preprocessing](03_preprocessing.qmd)** to transform,
normalise, and explore the QC-filtered dataset.
:::
