---
title: "Step 4 — Dimensionality Reduction"
subtitle: "Jackson & Fischer et al. (2020) Breast Cancer | imcdatasets"
author: "mass-cytometry-explorations"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: false
    code-tools: true
    fig-width: 9
    fig-height: 6
    embed-resources: true
conda:
  environment: mass-cytometry
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

## Overview

This notebook performs **Step 4** of the replication workflow: dimensionality
reduction on the preprocessed single-cell object for downstream clustering and
annotation.

Inputs from Step 3:

- `results/sce_preprocessed.rds`

Outputs from this notebook:

- `results/dimensionality_reduction_summary/dimensionality_reduction_parameters.csv`
- `results/dimensionality_reduction_summary/pca_variance_explained.csv`
- `results/dimensionality_reduction_summary/pca_embedding.csv`
- `results/dimensionality_reduction_summary/umap_embedding.csv` (if available)
- `results/dimensionality_reduction_summary/tsne_embedding.csv` (if available)
- `results/sce_dimensionality_reduced.rds`

---

## Setup

```{r setup}
suppressPackageStartupMessages({
  library(SingleCellExperiment)
  library(scater)
  library(ggplot2)
  library(dplyr)
  library(patchwork)
})

if (basename(getwd()) == "notebooks") {
  notebooks_dir <- normalizePath(getwd())
} else if (dir.exists(file.path(getwd(), "notebooks"))) {
  notebooks_dir <- normalizePath(file.path(getwd(), "notebooks"))
} else {
  stop("Run from project root or notebooks/ directory.")
}

if (!identical(normalizePath(getwd()), notebooks_dir)) {
  setwd(notebooks_dir)
}

cat("Working directory:", getwd(), "\n")
```

```{r paths}
dir.create(file.path("..", "results", "dimensionality_reduction_summary"),
           recursive = TRUE, showWarnings = FALSE)
dir.create(file.path("..", "figures", "04_dimensionality_reduction"),
           recursive = TRUE, showWarnings = FALSE)
```

---

## Load preprocessed data

```{r load-sce}
sce_path <- file.path("..", "results", "sce_preprocessed.rds")
if (!file.exists(sce_path)) {
  stop("Missing results/sce_preprocessed.rds. Please render notebooks/03_preprocessing.qmd first.")
}

sce_preprocessed <- readRDS(sce_path)
sce_preprocessed
```

---

## Configure DR feature set and assay

```{r dr-setup}
dr_assay <- if ("scaled" %in% assayNames(sce_preprocessed)) {
  "scaled"
} else if (!is.null(metadata(sce_preprocessed)$preprocessing$analysis_assay) &&
           metadata(sce_preprocessed)$preprocessing$analysis_assay %in% assayNames(sce_preprocessed)) {
  metadata(sce_preprocessed)$preprocessing$analysis_assay
} else {
  assayNames(sce_preprocessed)[1]
}

feature_flag_present <- "use_for_dr" %in% colnames(rowData(sce_preprocessed))
flagged_features <- if (feature_flag_present) {
  rownames(sce_preprocessed)[as.logical(rowData(sce_preprocessed)$use_for_dr)]
} else {
  character(0)
}

dr_features <- if (length(flagged_features) >= 2) {
  flagged_features
} else {
  rownames(sce_preprocessed)
}

n_pcs_target <- 20L
n_umap_pcs <- 15L
n_tsne_pcs <- 15L

cat("DR assay:", dr_assay, "\n")
cat("Number of features used for DR:", length(dr_features), "\n")
```

---

## Principal component analysis (PCA)

```{r run-pca}
mat <- t(assay(sce_preprocessed, dr_assay)[dr_features, , drop = FALSE])
max_rank <- max(2L, min(n_pcs_target, ncol(mat), nrow(mat) - 1L))

pca_fit <- prcomp(mat, center = TRUE, scale. = FALSE, rank. = max_rank)
pca_embed <- as.data.frame(pca_fit$x)

if (ncol(pca_embed) < 2) {
  stop("PCA produced fewer than 2 components; unable to proceed.")
}

reducedDim(sce_preprocessed, "PCA") <- as.matrix(pca_embed)

pca_var_explained <- (pca_fit$sdev^2) / sum(pca_fit$sdev^2) * 100
pca_var_tbl <- data.frame(
  component = seq_along(pca_var_explained),
  variance_explained_pct = pca_var_explained,
  cumulative_variance_explained_pct = cumsum(pca_var_explained)
)

cat(sprintf("Computed PCA with %d components.\n", ncol(pca_embed)))
```

```{r fig-pca-variance}
#| fig-cap: "PCA variance explained by component."
ggplot(pca_var_tbl, aes(x = component, y = variance_explained_pct)) +
  geom_col(fill = "steelblue", width = 0.8) +
  geom_line(aes(y = cumulative_variance_explained_pct), colour = "firebrick", linewidth = 0.8) +
  geom_point(aes(y = cumulative_variance_explained_pct), colour = "firebrick", size = 1.2) +
  labs(
    title = "PCA variance explained",
    subtitle = "Bars: per-component variance, red line: cumulative variance",
    x = "Principal component",
    y = "Variance explained (%)"
  ) +
  theme_bw()
```

```{r fig-pca-embedding}
#| fig-cap: "PCA embedding (PC1 vs PC2), coloured by log10 total counts."
cell_total_counts <- if ("counts" %in% assayNames(sce_preprocessed)) {
  colSums(assay(sce_preprocessed, "counts"), na.rm = TRUE)
} else {
  rowSums(mat, na.rm = TRUE)
}

pca_plot_df <- data.frame(
  PC1 = pca_embed$PC1,
  PC2 = pca_embed$PC2,
  log10_total_counts = log10(cell_total_counts + 1),
  check.names = FALSE
)

ggplot(pca_plot_df, aes(x = PC1, y = PC2, colour = log10_total_counts)) +
  geom_point(alpha = 0.5, size = 0.7) +
  scale_colour_viridis_c(option = "C") +
  labs(
    title = "PCA embedding",
    subtitle = sprintf("Assay: %s | Features: %d", dr_assay, length(dr_features)),
    x = "PC1",
    y = "PC2",
    colour = "log10(total counts + 1)"
  ) +
  theme_bw()
```

---

## Non-linear embeddings (UMAP and t-SNE)

```{r run-nonlinear}
sce_dr <- sce_preprocessed

umap_ok <- FALSE
tsne_ok <- FALSE
umap_error <- NA_character_
tsne_error <- NA_character_

umap_dims <- min(n_umap_pcs, ncol(reducedDim(sce_dr, "PCA")))
tsne_dims <- min(n_tsne_pcs, ncol(reducedDim(sce_dr, "PCA")))

try({
  sce_dr <- scater::runUMAP(
    sce_dr,
    dimred = "PCA",
    n_dimred = umap_dims,
    n_neighbors = 30,
    min_dist = 0.3,
    name = "UMAP"
  )
  umap_ok <- TRUE
}, silent = TRUE)

if (!umap_ok) {
  umap_error <- "UMAP failed (likely missing runtime dependency such as uwot)."
}

try({
  sce_dr <- scater::runTSNE(
    sce_dr,
    dimred = "PCA",
    n_dimred = tsne_dims,
    perplexity = 30,
    name = "TSNE"
  )
  tsne_ok <- TRUE
}, silent = TRUE)

if (!tsne_ok) {
  tsne_error <- "t-SNE failed (likely missing runtime dependency such as Rtsne)."
}

if (!umap_ok) cat("UMAP status:", umap_error, "\n")
if (!tsne_ok) cat("t-SNE status:", tsne_error, "\n")

cat("UMAP computed:", umap_ok, "\n")
cat("t-SNE computed:", tsne_ok, "\n")
```

```{r fig-nonlinear, eval=TRUE}
#| fig-cap: "UMAP/t-SNE embeddings when available."
plot_list <- list()

if (umap_ok) {
  umap_df <- as.data.frame(reducedDim(sce_dr, "UMAP"))
  plot_list[["UMAP"]] <- ggplot(
    umap_df,
    aes(x = UMAP1, y = UMAP2, colour = log10(cell_total_counts + 1))
  ) +
    geom_point(alpha = 0.5, size = 0.7) +
    scale_colour_viridis_c(option = "C") +
    labs(
      title = "UMAP embedding",
      x = "UMAP1",
      y = "UMAP2",
      colour = "log10(total counts + 1)"
    ) +
    theme_bw()
}

if (tsne_ok) {
  tsne_df <- as.data.frame(reducedDim(sce_dr, "TSNE"))
  plot_list[["tSNE"]] <- ggplot(
    tsne_df,
    aes(x = TSNE1, y = TSNE2, colour = log10(cell_total_counts + 1))
  ) +
    geom_point(alpha = 0.5, size = 0.7) +
    scale_colour_viridis_c(option = "C") +
    labs(
      title = "t-SNE embedding",
      x = "t-SNE1",
      y = "t-SNE2",
      colour = "log10(total counts + 1)"
    ) +
    theme_bw()
}

if (length(plot_list) == 0) {
  plot.new()
  text(0.5, 0.5, "UMAP/t-SNE not available in current runtime.")
} else if (length(plot_list) == 1) {
  plot_list[[1]]
} else {
  plot_list[[1]] + plot_list[[2]]
}
```

---

## Save outputs

```{r save-outputs}
params_tbl <- data.frame(
  parameter = c(
    "dr_assay",
    "n_cells",
    "n_markers",
    "n_features_used_for_dr",
    "n_pcs_computed",
    "umap_computed",
    "tsne_computed",
    "umap_pca_dims",
    "tsne_pca_dims"
  ),
  value = c(
    dr_assay,
    as.character(ncol(sce_dr)),
    as.character(nrow(sce_dr)),
    as.character(length(dr_features)),
    as.character(ncol(reducedDim(sce_dr, "PCA"))),
    as.character(umap_ok),
    as.character(tsne_ok),
    as.character(umap_dims),
    as.character(tsne_dims)
  )
)

write.csv(params_tbl,
          file = file.path("..", "results", "dimensionality_reduction_summary",
                           "dimensionality_reduction_parameters.csv"),
          row.names = FALSE)

write.csv(pca_var_tbl,
          file = file.path("..", "results", "dimensionality_reduction_summary",
                           "pca_variance_explained.csv"),
          row.names = FALSE)

pca_out <- data.frame(cell_id = colnames(sce_dr), reducedDim(sce_dr, "PCA"), check.names = FALSE)
write.csv(pca_out,
          file = file.path("..", "results", "dimensionality_reduction_summary",
                           "pca_embedding.csv"),
          row.names = FALSE)

if (umap_ok) {
  umap_out <- data.frame(cell_id = colnames(sce_dr), reducedDim(sce_dr, "UMAP"), check.names = FALSE)
  write.csv(umap_out,
            file = file.path("..", "results", "dimensionality_reduction_summary",
                             "umap_embedding.csv"),
            row.names = FALSE)
}

if (tsne_ok) {
  tsne_out <- data.frame(cell_id = colnames(sce_dr), reducedDim(sce_dr, "TSNE"), check.names = FALSE)
  write.csv(tsne_out,
            file = file.path("..", "results", "dimensionality_reduction_summary",
                             "tsne_embedding.csv"),
            row.names = FALSE)
}

saveRDS(sce_dr, file = file.path("..", "results", "sce_dimensionality_reduced.rds"))

cat("Saved:\n")
cat("  results/dimensionality_reduction_summary/dimensionality_reduction_parameters.csv\n")
cat("  results/dimensionality_reduction_summary/pca_variance_explained.csv\n")
cat("  results/dimensionality_reduction_summary/pca_embedding.csv\n")
if (umap_ok) cat("  results/dimensionality_reduction_summary/umap_embedding.csv\n")
if (tsne_ok) cat("  results/dimensionality_reduction_summary/tsne_embedding.csv\n")
cat("  results/sce_dimensionality_reduced.rds\n")
```

---

## Session information

```{r session-info}
sessionInfo()
```

---

::: {.callout-tip}
## Next step

Proceed to clustering and cluster annotation using
`results/sce_dimensionality_reduced.rds` as input.
:::
